---
title: "Coverage of holistic needs assessments and personalised care and support plans - England diagnoses in 2021"
format: revealjs
editor: visual
embed-resources: true
---

```{r prep}
library(kableExtra)
library(scales)
library(gridExtra)
library(tidyverse)

load("All_data.RData")
```

## Methods

```{r overall_figures}
percent_unique_all <- (unique_hnas_pcsps/all_records)*100
percent_dups_all <- 100-percent_unique_all
percent_pts_with_hna <- (pts_with_hna/hnas_denom_size)*100
percent_pts_with_pcsp <- (pts_with_pcsp/pcsps_denom_size)*100
percent_only_hna <- (only_hna/ids_either)*100
percent_only_pcsp <- (only_pcsp/ids_either)*100
percent_both <- (both/ids_either)*100
percent_notreqpcsps <- (pcsps_notreq/unique_off_pcsps)*100
```

## Overall

There were `r all_records` of a HNA or PCSP in the RCRD table associated with someone diagnosed in 2021, occurring within 2 years of diagnosis. Of these, `r percent_unique_all`% (`r unique_hnas_pcsps`) were unique records, meaning `r percent_dups_all`% were duplicates. This is believed to be primarily because of the way COSD data is submitted by Trusts. Each time something changes in a patient's record, the entire COSD record is resubmitted, meaning a duplicate of any previous HNAs and PCSPs will be created. 

There were `r unique_hnas` unique HNAs recorded for patients diagnosed in 2021 and `r unique_pcsps` unique PCSPs.

`r hnas_notoff` unique HNAs and `r pcsps_notoff` unique PCSPs were recorded as 'not offered', meaning there were `r unique_off_hnas` unique HNAs and `r unique_off_pcsps` unique PCSPs which were offered to patients diagnosed in 2021 within 2 years of their diagnosis. This includes `r pcsps_notreq` recorded as 'not required (no concerns from HNA)'. 

#Status:
Excluding those recorded as 'not offered', the status codes recorded against HNAs and PCSPs are as follows:

```{r number_records_table}

offered_code_data <- offered_code_data %>%
  select(-count, -total_count) %>%
  pivot_wider(., names_from = event_type, values_from = percent) %>%
  mutate(HNA = ifelse(is.na(HNA), "-", HNA)) %>%
  rename("Status" = "offered_code")

kable(offered_code_data, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) %>%
  add_header_above(c(" " = 1, "Proportion of patients" = 2))
  
```

#Patients:
The initial denominator, using the approach described above, was `r denom_size` patients, compared to 329,665 reported in the national statistics for 2021. Some Trusts are not submitting HNA/PCSP data to COSD, meaning patients diagnosed at these Trusts should be excluded from coverage analysis. Trusts with fewer than 20 HNAs/PCSPs recorded were excluded. This left a denominator of `r hnas_denom_size` patients for HNAs and `r pcsps_denom_size` patients for PCSPs.

Of these:
-   `r percent_pts_with_hna`% of patients had at least one record of being offered a HNA within 2 years of diagnosis (`r pts_with_hna` patients)
-   `r percent_pts_with_pcsp`% of patients had at least one record of being offered a PCSP (or a PCSP being deemed 'not required') within 2 years of diagnosis (`r pts_with_pcsp` patients)

Of the patients who had at least one record of either a HNA or PCSP (`r ids_either`):
- `r percent_only_hna`% had only a HNA recorded (`r only_hna`)
- `r percent_only_pcsp`% had only a PCSP recorded (`r only_pcsp`)
- `r percent_both`% had both a HNA and PCSP recorded (`r both`)

Some patients have more than one HNA/PCSP:

```{r number_records_table}

count_summary_table <- left_join(hna_count_summary, pcsp_count_summary, by = c("hna_count" = "pcsp_count"))

count_summary_table <- count_summary_table %>%
  rename("Number of unique HNA/PCSP records" = "hna_count") %>%
  rename("HNAs" = "percent_patients_hna") %>%
  rename("PCSPs" = "percent_patients_pcsp") %>%
  select(-c(number_patients_hna, number_patients_pcsp)) 

kable(count_summary_table, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) %>%
  add_header_above(c(" " = 1, "Proportion of patients" = 2))
  
```

## Completeness and quality of COSD data

After excluding the non-submitting Trusts from the data, there were `r unique_hnas` unique HNAs and `r unique_pcsps` unique PCSPs remaining in the dataset. 

```{r completeness_table}

kable(cosd_completeness, align = "llcc", col.names = c("Quality measure", 
                                                     "Type of event", 
                                                     "Number of unique records", 
                                                     "Percentage of unique records")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 
  
  
```


## Comparison to eHNA data
## Delivery across the pathway
inc "hnas and pcsps"

Of those unique HNA/PCSP records (other than those which were 'Not offered') with a valid point in pathway code: 
```{r point_in_pathway}

pathway_data <- pathway_data %>%
  select(-count, -total_count) %>%
  pivot_wider(., names_from = "event_type", values_from = "percent") 

kable(pathway_data, align = "lcc", col.names = c("Point in pathway", 
                                                 "Proportion of HNAs", 
                                                 "Proportion of PCSPs")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 

pathway_graph
  
```

## Staff roles

Of those unique HNA/PCSP records (other than those which were 'Not offered') with a valid staff role code: 
```{r staff_roles}

staff_role_data <- staff_role_data %>%
  select(-count, -total_count) %>%
  pivot_wider(., names_from = "event_type", values_from = "percent") 

kable(staff_role_data, align = "lcc", col.names = c("Staff role", 
                                                    "Proportion of HNAs", 
                                                    "Proportion of PCSPs")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 

staff_role_graph
  
```

## Inequalities
```{r gender}

gender_hna %>% select(-lower, -upper, -percent) %>%
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Gender", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 

gender_pcsp %>% select(-lower, -upper, -percent) %>%
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Gender", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 


grid.arrange(gender_hna_graph, gender_pcsp_graph, ncol=2)

```

```{r age}

age_hna %>% select(-lower, -upper, -percent) %>%
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 

age_pcsp %>% select(-lower, -upper, -percent) %>%
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 


grid.arrange(age_hna_graph, age_pcsp_graph, ncol=2)

```

```{r ethnicity}

ethnicity_hna %>% select(-lower, -upper, -percent) %>%
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 

ethnicity_pcsp %>% select(-lower, -upper, -percent) %>%
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 10) 


grid.arrange(ethnicity_hna_graph, ethnicity_pcsp_graph, ncol=2)

```
## Geographic variation
