---
title: "Coverage of holistic needs assessments and personalised care and support plans - England diagnoses in 2021"
format: 
  revealjs:
    smaller: true
    scrollable: true
    footer: "For internal use only"
    logo: "Logo.png"
    width: 1000
    css: styles.css
editor: visual
execute:
  echo: false
  warning: false
  ####do a css styles file
embed-resources: true
---

```{r prep}
library(kableExtra)
library(scales)
library(gridExtra)
library(tidyverse)

load("All_data.RData")
```

## 1. Background

Personalised care and support planning is a process to ensure patients' physical, practical, emotional and social needs are identified and supported. Guidance on using holistic needs assessment (HNA) to deliver personalised care was first published in 2007 and since then HNAs have been rolled out in the NHS for patients diagnosed with cancer (1-3). The NHS Long Term Plan states that "where appropriate every person diagnosed with cancer will have access to personalised care, including needs assessment, a care plan and health and wellbeing information and support" (4).

HNAs use a structured questionnaire approach to gather information about patients' physical, social, emotional and financial needs, as well as their psychological and spiritual wellbeing (1-3). HNAs are delivered using a range of tools including the Macmillan eHNA system (2, 3) and are used to identify patients' concerns and develop a Personalised Care and Support Plan (PCSP) (2, 3).

Since 2020 and the launch of the NHS England Cancer Outcomes and Services Dataset (COSD) v9.0, more information on HNAs and PCSPs delivered at all NHS Trusts has been collected. NHS organisations are required to report data on the offer of at least one HNA for every patient diagnosed with cancer, using four data items: assessment offered, assessment completed date, assessment point of pathway, and staff role carrying out the assessment. In May 2021, this was expanded to mandate recording of the same data for PCSPs. Unlike the Macmillan eHNA data, these fields do not provide information on the needs that patients highlighted as part of the assessment, however they could still provide useful insights into the characteristics of those who are offered and who complete HNAs and PCSPs. Most NHS Trusts are now submitting HNA and PCSP data, meaning we can begin to explore the delivery of HNAs and PCSPs.

There is currently some research exploring the efficacy of HNAs, methods for using them to effectively capture patient concerns, and the mechanisms by which assessments might influence self-advocacy (5-8). However, there is minimal evidence about inequalities in the delivery of HNAs, PCSPs, and personalised care more generally. Understanding the coverage of HNAs and PCSPs will help to evaluate the delivery of personalised care for people living with cancer and inform the provision of better personalised care across the NHS. Macmillan are interested in understanding the coverage of HNAs to inform their delivery of the Macmillan eHNA platform and wider activity in personalised care.

## 2. Methods

Data on HNAs and PCSPs are submitted to the NHS England Cancer Outcomes and Services Dataset (COSD). Those records with an associated date are available in the Rapid Cancer Registration Dataset (RCRD). RCRD can be linked on NHS number to patient diagnosis records.

Patient cohort:

Data on patients diagnosed in 2021 with a malignant cancer (C00-C97 excluding C44) were extracted from the registry. For patients with multiple tumours diagnosed in 2021, their first diagnosed tumour is used. All patients are included, regardless of whether they have any subsequent or previous diagnoses. Patients diagnosed with a stage 0 tumour were excluded from the analysis. Also excluded were death certificate only diagnoses, and patients who were diagnosed and died on the same day. There was a small number of patients with a death date recorded before their diagnosis date, who were also excluded.

Some trusts are not submitting HNA and PCSP data to COSD, meaning patients diagnosed at these trusts were excluded from this analysis. Patients diagnosed at trusts meeting the following criteria were excluded:

-   Trusts with fewer than 10 HNA/PCSP records and fewer than 100 patients diagnosed in 2021

-   Trusts with fewer than 20 HNA/PCSP records and at least 100 patients diagnosed in 2021

This left a denominator of `r scales::comma(hnas_denom_size)` patients for HNA coverage analysis and `r scales::comma(pcsps_denom_size)` for PCSP coverage analysis, compared to 329,665 patients diagnosed in 2021 recorded in national statistics.

HNA and PCSP records:

HNA and PCSP records in RCRD were linked to the patient cohort on NHS number. There were `r scales::comma(all_records)` of a HNA or PCSP in the RCRD table associated with someone diagnosed in 2021, occurring within 2 years of diagnosis. Of these, `r scales::percent(percent_unique_all, accuracy = 0.01)`% (`r scales::comma(unique_hnas_pcsps)`) were unique records, meaning `r scales::percent(percent_dups_all, accuracy = 0.01)`% were duplicates. This is believed to be primarily because of the way COSD data is submitted by Trusts. Each time something changes in a patient's record, the entire COSD record is resubmitted, meaning a duplicate of any previous HNAs and PCSPs will be created.

Duplicates were removed and unique HNA and PCSP records associated with each patients were linked to the patient cohort. All HNAs and PCSPs occurring within 2 years of the patient's diagnosis date were assessed and the earliest HNA record identified for each patient. The analysis primarily explores the first HNA and PCSP for each patient.

```{r overall_figures}
percent_unique_all <- (unique_hnas_pcsps/all_records)
percent_dups_all <- 1-percent_unique_all
percent_pts_with_hna <- (pts_with_hna/hnas_denom_size)
percent_pts_with_pcsp <- (pts_with_pcsp/pcsps_denom_size)
percent_only_hna <- (only_hna/ids_either)
percent_only_pcsp <- (only_pcsp/ids_either)
percent_both <- (both/ids_either)
percent_notreqpcsps <- (pcsps_notreq/unique_off_pcsps)
```

## 3. Status of HNA and PCSP records

There were `r scales::comma(unique_hnas)` unique HNAs recorded for patients diagnosed in 2021 and `r scales::comma(unique_pcsps)` unique PCSPs.

`r scales::comma(hnas_notoff)` unique HNAs and `r scales::comma(pcsps_notoff)` unique PCSPs were recorded as 'not offered', meaning there were `r scales::comma(unique_off_hnas)` unique HNAs and `r scales::comma(unique_off_pcsps)` unique PCSPs which were offered to patients diagnosed in 2021 within 2 years of their diagnosis. This includes `r scales::comma(pcsps_notreq)` recorded as 'not required (no concerns from HNA)'.

###### Table 3a: Number and proportion of HNA and PCSP records with each status code, all unique records for patients diagnosed in 2021, excluding those recorded as 'not offered'

```{r offered_code_table, out.width='100%', fig.align='center'}

offered_code_data <- offered_code_data |>
  select(-c(count, total_count, percent_graph)) |>
  pivot_wider(names_from = event_type, values_from = percent) |>
  mutate(HNA = ifelse(is.na(HNA), "-", HNA)) |>
  rename("Status" = "offered_code")

kable(offered_code_data, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") |>
  add_header_above(c(" " = 1, "Proportion of records" = 2))
  
```

## 4. Completeness and quality of COSD data

After excluding the non-submitting Trusts from the data, there were `r scales::comma(unique_hnas)` unique HNAs and `r scales::comma(unique_pcsps)` unique PCSPs remaining in the dataset.

COSD HNA and PCSP data includes variables for the 'point in pathway' when the HNA/PCSP was delivered, and the staff role of the person delivering the HNA/PCSP. There is some missing data in these fields (Table 4a). Some records are also recorded as having occurred on dates after the patient's recorded date of death (Table 4a).

###### Table 4a: Number and proportion of HNA and PCSP records with each quality measure characteristic, all unique records for patients diagnosed in 2021

```{r completeness_table, out.width='100%', fig.align='center'}

kable(cosd_completeness, align = "lccc", col.names = c("Quality measure",                                                       "Type of event",                                                       "Number of records",                                                       "Proportion of records"), format.args = list(big.mark = ",")) |>   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center")

```

## 5. Comparison to eHNA data

Data from the eHNA system for assessments marked as "locked" or "submitted", recorded by each NHS Trust, were compared to the number of unique HNAs recorded as "offered and accepted" in RCRD by Trust, occurring in 2021. This COSD HNA dataset contains HNAs occurring in 2021, rather than associated with patients diagnosed in 2021, as used in the other parts of this analysis.

There are significantly more HNAs recorded in the eHNA data in 2021, than in COSD. `r length(not_in_ehna)` Trusts with data in RCRD did not appear in the eHNA data. `r length(not_in_cosd)` Trusts in the eHNA data did not appear in RCRD. Of those Trusts which do appear in both datasets, `r more_in_cosd` have more HNAs recorded in COSD than eHNA and `r more_in_ehna` have more recorded in eHNA than COSD.

###### Table 5a: Difference in the number of unique HNA records in COSD compared to eHNA by NHS Trust in 2021

```{r comparison_table, out.width='100%', fig.align='center'}  differences_graph}
```

## 

## 6. Overall coverage

`r scales::percent(percent_pts_with_hna, accuracy = 0.01)`% of patients had at least one record of being offered a HNA within 2 years of diagnosis (`r scales::comma(pts_with_hna)` patients)

`r scales::percent(percent_pts_with_pcsp, accuracy = 0.01)`% of patients had at least one record of being offered a PCSP (or a PCSP being deemed 'not required') within 2 years of diagnosis (`r scales::comma(pts_with_pcsp)` patients)

## 7. Delivery across the pathway

Combinations:

Of the patients who had at least one record of either a HNA or PCSP (`r scales::comma(ids_either)`): - `r scales::percent(percent_only_hna, accuracy = 0.01)`% had only a HNA recorded (`r scales::comma(only_hna)`) - `r scales::percent(percent_only_pcsp, accuracy = 0.01)`% had only a PCSP recorded (`r scales::comma(only_pcsp)`) - `r scales::percent(percent_both, accuracy = 0.01)`% had both a HNA and PCSP recorded (`r scales::comma(both)`)

Multiple HNAs and PCSPs:

###### Table 7a: Proportion of patients diagnosed in 2021 receiving each number of HNAs and PCSPs in the two years post-diagnosis 

```{r number_records_table, out.width='100%', fig.align='center'}  count_summary_table <- left_join(hna_count_summary, pcsp_count_summary, by = c("hna_count" = "pcsp_count"))  count_summary_table <- count_summary_table |>   rename("Number of unique HNA/PCSP records" = "hna_count") |>   rename("HNAs" = "percent_patients_hna") |>   rename("PCSPs" = "percent_patients_pcsp") |>   select(-c(number_patients_hna, number_patients_pcsp))   kable(count_summary_table, align = "c") |>   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") |>   add_header_above(c(" " = 1, "Proportion of patients" = 2))}
```

Point in pathway code:

COSD collects the 'point in pathway' when a HNA/PCSP is delivered. This is defined by the person submitting the record using a numerical code.

###### Table 7b: Proportion of HNA and PCSP records with each point of pathway code, all unique records for patients diagnosed in 2021, excluding those recorded as 'not offered'

```{r point_in_pathway, out.width='100%', fig.align='center'}

pathway_data <- pathway_data |>
  select(-c(count, total_count, percent_graph)) |>
  pivot_wider(names_from = "event_type", values_from = "percent") 

kable(pathway_data, align = "lcc", col.names = c("Point in pathway", 
                                                 "Proportion of HNAs", 
                                                 "Proportion of PCSPs")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

pathway_graph
  
```

## 8. Staff roles

COSD collects the 'staff role' of the person delivering a HNA/PCSP. This is defined by the person submitting the record using a numerical code.

###### Table 8a: Proportion of HNA and PCSP records with each staff role code, all unique records for patients diagnosed in 2021, excluding those recorded as 'not offered'

```{r staff_roles, out.width='100%', fig.align='center'}

staff_role_data <- staff_role_data |>
  select(-c(count, total_count, percent_graph)) |>
  pivot_wider(names_from = "event_type", values_from = "percent") 

kable(staff_role_data, align = "lcc", col.names = c("Staff role", 
                                                    "Proportion of HNAs", 
                                                    "Proportion of PCSPs")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

staff_role_graph
  
```

## Variation in coverage - gender

```{r gender, out.width='100%', fig.align='center'}

gender_hna |> select(-lower, -upper, -percent) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Gender", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

gender_pcsp |> select(-lower, -upper, -percent) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Gender", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 


grid.arrange(gender_hna_graph, gender_pcsp_graph, ncol=2)

```

## Variation in coverage - age

```{r age, out.width='100%'}

age_hna |> select(-lower, -upper, -percent) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

age_pcsp |> select(-lower, -upper, -percent) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 


grid.arrange(age_hna_graph, age_pcsp_graph, ncol=2)

```

## Variation in coverage - ethnicity

```{r ethnicity, out.width='100%', fig.align='center'}

ethnicity_hna |> select(-lower, -upper, -percent) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

ethnicity_pcsp |> select(-lower, -upper, -percent) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Age group", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 


grid.arrange(ethnicity_hna_graph, ethnicity_pcsp_graph, ncol=2)

```

## Variation in coverage - deprivation

```{r imd, out.width='100%', fig.align='center'}

imd_hna |> select(-c(lower, upper, percent)) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Deprivation decile", "HNA status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

imd_pcsp |> select(-c(lower, upper, percent)) |>
  kable(format.args = list(big.mark = ","), align = "llcccc", 
        col.names = c("Deprivation decile", "PCSP status", "Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 


grid.arrange(imd_hna_graph, imd_pcsp_graph, ncol=2)

```

## Variation in coverage - tumour type

```{r site, out.width='100%', fig.align='center'}

site_hna |> select(-c(lower, upper, percent, hna_status)) |>
  kable(format.args = list(big.mark = ","), align = "lcccc", 
        col.names = c("Tumour type","Number of patients", "Proportion of patients",
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

site_pcsp |> select(-c(lower, upper, percent, pcsp_status)) |>
  kable(format.args = list(big.mark = ","), align = "lcccc", 
        col.names = c("Tumour type","Number of patients", "Proportion of patients",
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 


grid.arrange(site_hna_graph, site_pcsp_graph, ncol=2)
```

## Variation in coverage - stage at diagnosis

```{r stage, out.width='100%', fig.align='center'}

stage_hna |> select(-c(lower, upper, percent, hna_status)) |>
  kable(format.args = list(big.mark = ","), align = "lcccc", 
        col.names = c("Stage at diagnosis","Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

stage_pcsp |> select(-c(lower, upper, percent)) |>
  kable(format.args = list(big.mark = ","), align = "lcccc", 
        col.names = c("Stage at diagnosis","Number of patients", "Proportion of patients", 
                      "Lower 95% confidence limit", "Upper 95% confidence limit")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 


grid.arrange(stage_hna_graph, stage_pcsp_graph, ncol=2)
```

## Variation in coverage - geography

Geographic variation is measured by the trust of diagnosis, which is assumed to be the most likely trust delivering first HNAs and PCSPs.

```{r geog, out.width='100%', fig.align='center'}

trust_coverage_hna_cats |> 
  kable(format.args = list(big.mark = ","), align = "lc", 
        col.names = c("Proportion of patients offered a HNA", "Number of trusts")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

trust_coverage_pcsp_cats |> 
  kable(format.args = list(big.mark = ","), align = "lc", 
        col.names = c("Proportion of patients offered a PCSP", "Number of trusts")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 14, position = "center") 

grid.arrange(trust_hna_graph, trust_pcsp_graph, ncol=2)

```

NHS trusts do not have clearly defined catchment areas, so the below shows the average coverage in NHS trusts within each ICB.
